### 2025-09-18 paper/sec/introduction.tex
Rehashing para 3.

Making sense of and verifying data-driven claims is hard, even when the evidence base lives in the public
domain. A key challenge lies in tracing specific claims back to the underlying data. In peer review, for
example, empirical claims typically lack author-supplied links to data, making it hard for reviewers to check
them directly~\citep{weber20}. Paper retractions, meanwhile, are often attributable not to fraud, but to
simple errors in data management or analysis~\citep{hu25}. Large language models (LLMs) can assist with
fact-checking~\citep{abu-ahmad25} and interpretation of charts and figures~\citep{roberts24}, but current LLM
interfaces do not support direct interrogation of visual or other outputs.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25}, on the other
hand, takes a more infrastructural approach to the transparency problem, by linking visualisations to their
data sources directly. This allows visual outputs to support in situ \emph{provenance queries}, user
interactions (e.g. mousing over a visual element) that reveal how visual features relate to data. The value
proposition of these approaches is that the interactions are provided automatically via trusted
infrastructure, namely a query language or general-purpose programming language which tracks dependencies as
data flows through a computation. However, these approaches to transparency only support direct interrogation
of outputs computed from data, such as visualisations. The accompanying natural language, which often plays a
central role in making the claims of the paper, remains hard to interpret.

In this paper, we make the case for combining these two approaches, leveraging the ability of LLMs to
understand complex technical natural language and to synthesise programs that express the analyses that
underwrite those claims, and extend interactive provenance queries to the natural language that accompanies
the charts and figures.  ..language which also makes quantitative statements and contains substantial implicit
analytic content. In principle transparent programming languages can also be used to create natural language
which is bidirectly linked to data, by computing key fragments of natural language from data. The following
example, adapted from \cite{zhang18} via \cite{moosavi21}, shows how the language presented in this paper can
be used to achieve this \rpnote{Need to illustrate counterfactual/generalisability}:

>>>>

Making sense of and verifying data-driven claims is hard, even when the evidence base lives in the public
domain. A key challenge lies in tracing specific claims back to the underlying data. In peer review, for
example, empirical claims typically lack author-supplied links to data, making them hard for reviewers to
check directly~\citep{weber20}. Paper retractions, meanwhile, are often attributable not to fraud, but to
simple errors in data management or analysis~\citep{hu25}. Large language models (LLMs) can assist with
fact-checking~\citep{abu-ahmad25} and interpretation of charts and figures~\citep{roberts24}, but current LLM
interfaces do not support direct interrogation of visual or other outputs.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25}, on the other
hand, takes a more infrastructural approach to the transparency problem, linking visualisations to their data
sources directly. This allows visual outputs to support \emph{provenance queries}, user interactions (e.g.
mousing over a visual element) that reveal how visual features relate to data. The advantage of this approach
is that the interactions are provided automatically via trusted infrastructure, namely a query language or
general-purpose programming language which tracks dependencies as data flows through a computation. However,
these approaches are limited to outputs computed from data, such as visualisations. What is missing is a way
to extend these ``direct interrogation'' features to natural language itself, where the main claims of most
scholarly articles are actually made.

In this paper, we make the case that progress can be made on this problem by combining these two approaches.
We introduce the idea of \emph{transparent documents}, web-based scholarly articles where key fragments of
natural language are computed from data, like visualisations and other summaries, and that support interactive
provenance queries that reveal how key quantitative statements in the document relate to the data. We report
on a proof-of-concept LLM-based tool which supports a human-in-the-loop workflow for iteratively transforming
a preexisting opaque document into a transparent, data-driven counterpart. Our approach leverages the ability
of LLMs to understand technical language and synthesise code, and also the provenance-tracking runtime of an
open source programming language called Fluid, which we use as our target platform.

The following example, adapted from \cite{zhang18} via \cite{moosavi21}, shows how the language presented in
this paper can be used to achieve this \rpnote{Need to illustrate counterfactual/generalisability}:
